---
title: "Investigating Land and Freshwater Mollusc Diversity and Distribution in Gran Canaria, Spain"
subtitle: "Project description and planning"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-location: left
bibliography: [references.bib]
csl: american-psychological-association-7th-edition.csl
---

```{r, include=FALSE}
library(knitr)
library(here)
opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(root.dir = here::here())

library(kableExtra)
```

# Introduction {#sec-intro}

## Background {#sec-background}

Canary Islands (Spain) is an archipelago off the coast of northwest Africa. Gran Canaria is the third-largest island and located in the centre of the archipelago [@troll2016geology]. The island contains a number of endemic land snail species [e.g. @nunez2010silvestres]. Freshwater molluscs are of lesser significance, but are also present. Research on land and freshwater molluscs of the Canaries mainly goes back to the 19th century with the famous works of @webb1833synopsis, Shuttleworth [-@shuttleworth1852diagnoseneiniger; -@shuttleworth1852diagnosenneuer], @mousson1873revision, @wollaston1878testacea, @mabille1884materiaux, @odhner1931beitrage, and others. Recent checklists are available [@groh1985landschnecken; @bank2002catalogue; @nunez2010silvestres] (see also <https://www.malacowiki.org/>) as well as a limited number of more recent papers on specific taxa. Nevertheless, taxonomic research is still largely based on these old works and many species have never been found again since their description, or the ecology or proper range is not known. This, together with the threats of global warming [@luque2013temperature] and the increase of demographic and touristic pressure [@ibanez1997distribution], could have (and probably already has) detrimental consequences for the survival of these species on the island (see also the assessments of IUCN Red List of Threatened Species (<https://www.iucnredlist.org/>)). Therefore, there is an urgent need for information on ecology, distribution and taxonomy. We hope that this dataset of land and freshwater snail occurrences can contribute to the knowledge on these species and ensure their survival on Gran Canaria and the Canary Islands as a whole.

## Objectives {#sec-objectives}

The primary objective of this research project is to investigate the diversity and distribution of land and freshwater molluscs in Gran Canaria. Specifically, the project aims to:

1.  Conduct a literature study on the diversity and distribution of land and freshwater molluscs of Gran Canaria
2.  Identify the species present in Gran Canaria
3.  Determine the distribution of each species and identify gaps in current knowledge
4.  Provide a template for similar research in the other Canary Islands

## Methodology {#sec-methods}

In this manuscript we provide an overview and timeline of the work packages and tasks needed to complete the objectives described above. We identify the following work packages (WPs).

WP1: Communication and networking\
WP2: Literature study\
WP3: Data management\
WP4: Data processing and synthesis

These will be split in distinct tasks with a chronological order. No absolute time management will be made as this is merely a hobby. Finishing the tasks in chronological order will however improve progress of the project throughout time (@tbl-worktab).

```{r}
#| label: tbl-worktab
#| tbl-cap: "Overview of work packages and tasks" 

tasks_per_wp <- c(3, 5, 4, 8)
wp_df <- data.frame(
  work_package = rep(c("WP1", "WP2", "WP3", "WP4"), tasks_per_wp),
  tasknr = paste0("Task ", rep(1:length(tasks_per_wp), tasks_per_wp), 
                  ".", unlist(lapply(tasks_per_wp, seq_len))),
  task = c("Create questionaire for contributers and interested parties", 
           "Make dataset of contributer details", 
           "List of expected and possible outcomes",
           "Search for relevant sources",
           "Organize the sources", 
           "Evaluate and synthesize the sources",
           "Write comprehensive literature review",
           "Write literature report",
           "photos", "occurrence data", "data management plan", "data quality",
           "Use version control for reproducibility",
           "species list", "synonyms", "morphology", "ecology", "literature", 
           "photos", "distribution")
)

wp_df |>
  kable(col.names = c("Work package", "Task", "")) |>
  kable_classic(full_width = F, html_font = "Cambria")
```

# WP1: Communication and networking {#sec-WP1}

A lot of data and knowledge is in hands of amateur malacologists and private collections. Therefore, it is useful to know who can and wants to contribute to this project. This could be for example by sharing of data, literature, review of reports/papers etc. Even for people who are currently not interested to contribute or who might be interested in the future, it would be useful to have contact details. Obtaining correct contact details of everyone is also important so that everyone gets the appropriate recognition for the work they provided.

## Task 1.1: Create questionaire for contributers and interested parties {#sec-task11}

We would create a short questionnaire asking people to provide their basic contact details: `name`, `e-mail`, `adress` (optional) and `telephone number` (optional). In a second part we can ask how they want to stay informed on the project: `not at all`, `I want to be informed about interesting discoveries and intermediate results`. Thirdly, we can ask which contribution role they would like to play: `no role`, `contribute data`, `review reports/papers written in the light of this project`, `discuss and exchange literature`. In the end it must be clarified that a contribution in any for is very much appreciated and that everyone gets the appropriate recognition for this. It is also possible to be a coauthor.\
The questionaire will be created with [Google Forms](https://www.google.com/intl/en/forms/about/). We will ensure that the data collected from users is used appropriately, and their privacy and confidentiality are respected. Therefore we need to be upfront about how the data will be used and who will have access to it. Make sure users understand what information they are sharing, why it's being collected, and how it will be used. Make sure users give their informed consent to share their data, and clearly communicate how they can opt-out if they change their mind.

## Task 1.2 Make dataset of contributer details {#sec-task12}

To keep of track of everyone who has contributed in the past or wants to contribute in the future, the data from [task 1.1](#sec-task11) should be stored in a clean dataset. This to ensure that everyone gets the recognition they deserve even if it happened a long time ago.  
The dataset will be created with [Google Sheets](https://www.google.com/sheets/about/).

## Task 1.3 List of expected and possible outcomes {#sec-task13}

This task is performed here. There are three categories of possible outputs for this project.

1. Separate scientific papers
2. Full scale report
3. Website

### 1. Separate scientific papers

Papers on specific discoveries or topics can be published in malacological journals. These would be mainly ancillary papers related to the project, but not essential to the main goals. [Elona](https://elona-journal-malacology.blogspot.com/) is a digital journal on marine and continental Malacology in the Iberian Peninsula. The [Iberus](https://www.soesma.es/publicaciones/) journal of the Sociedad Española de Malacología and the Instituto Português de Malacologia. Any other non-Spanish journal is also an option (e.g. [Basteria](https://www.spirula.nl/basteria-engels/)).  
These papers will be written according to the guidelines of the intended journal.

### 2. Full scale report

A comprehensive report would be the main output of the project. More specifically, there would be two separate reports.

1. Literature study (@sec-WP2)
2. Species descriptions (@sec-WP4)

These will be written as HTML [Quarto Books](https://quarto.org/docs/books/) in RStudio [@RStudio2022]. This creates a very flexible output format for future adaptations especially in combination with version control (@sec-WP4).  
These reports could be published as books, scientific papers or on a website (see further).

### 3. Website

Essentially, HTML books are a specific form of [Quarto Websites](https://quarto.org/docs/websites/), meaning that they have the same capabilities as regular websites. There are several possibilities to publish Quarto Websites (<https://quarto.org/docs/publishing/>). A website could therefore easily contain the reports as well as other pages related to methodology (e.g. creation of UTM grids), extra material (e.g. extra pictures), background, blog ... of the project. Furthermore, the website could be updated once new data, literature ... comes available. In turn, this could entice additional individuals to participate in the project.

# WP2: Literature study {#sec-WP2}

We perform a critical analysis and synthesis of academic sources on the land and freshwater molluscs of Gran Canaria.  The main goal is to publish an overview of all available literature but also summarize these papers. This will in turn inform the research questions and methodology of this project.

## Task 2.1: Search for relevant sources {#sec-task21}

We will conduct a comprehensive search for academic sources, including books, journal articles, conference proceedings, and other relevant publications through a variety of databases (e.g. [Biodiversity Heritage Library](https://www.biodiversitylibrary.org/)), search engines (e.g. [Google Scholar](https://scholar.google.com/)), and keywords to ensure that you have a wide range of sources. A collection of usefull references related to the Macaronesian malacofauna can be found in @hemmen1989bibliographia and @bank2002catalogue.

## Task 2.2: Organize the sources {#sec-task22}

We make a dataset in Google Sheets of all sources. The first column `key` contains a unique identifier which will also be the key for citing this reference. The next columns are related to the [BibTeX](https://en.wikipedia.org/wiki/BibTeX) reference management software for formatting lists of references. First a column `type` which specifies the entry type (e.g. book, article, incollection, misc). The other columns are the fields related to the respective entry type (e.g. author, year, url, publisher, note).  
We write a script to convert this dataset to a .bib file to use for citing in texts. All analyses are done in with R in RStudio [@R2022; @RStudio2022] as [Quarto Documents](https://quarto.org/docs/output-formats/html-basics.html) (see methodology @sec-WP4). There are possibly already packages which can do this (e.g. [tbeptools](https://tbep-tech.github.io/tbeptools/reference/anlz_refs.html)).

## Task 2.3: Evaluate and synthesize the sources  {#sec-task23}

We make a dataset in Google Sheets of all sources. The first column `key` contains a unique identifier which will also be the key for citing this reference. The next columns are Boolean fields (True/False) group the sources according to common themes or topics.  
We distinguish the following categories of scientific sources:  
  
Empirical studies: These papers report on original research that involves collecting and analyzing data from field or laboratory experiments, observations, or surveys. They often include statistical analysis to test hypotheses or explore relationships between variables.  
  
Review papers: These papers provide a critical overview of the current state of knowledge on a particular ecological topic or research question. They typically summarize and synthesize findings from multiple studies and identify gaps, controversies, and future directions for research.    
  
Theoretical papers: These papers develop and test conceptual models, frameworks, or hypotheses related to ecological processes, interactions, or patterns. They often use mathematical or simulation models to explore theoretical scenarios or make predictions.  
  
Methodological papers: These papers describe and evaluate new or improved methods, techniques, or tools for ecological research, data analysis, or monitoring. They may involve field or laboratory experiments, statistical analyses, or software development.  
  
Applied papers: These papers focus on the practical application of ecological research to real-world problems, such as conservation, management, policy, or restoration. They often involve interdisciplinary collaboration and the integration of ecological, social, and economic considerations.  
  
Data papers: Peer reviewed documents that describe a particular dataset or collection of datasets. The primary purpose of a data paper is to provide a comprehensive description of the data to facilitate their reuse and replication by other researchers.  
  
Historical papers: These papers explore the historical context, evolution, and impact of ecological ideas, theories, or research methods. They may involve analysis of archival data, literature review, or oral history interviews.  
  
Species checklists: These papers provide a comprehensive list of species found in a particular geographic region, habitat, or taxonomic group. They often include information on the scientific and common names of the species, their ecological characteristics, and their conservation status. 
  
Unpublished sources: These papers include reports, theses, dissertations, or other research outputs that have not been formally published in peer-reviewed journals or books. They may be available in institutional repositories, conference proceedings, or other online platforms.
  
Grey literature: The body of documents and materials that are produced outside of traditional publishing channels and are not commercially published. This includes reports, theses, dissertations, conference proceedings, working papers, government documents, and other types of unpublished or non-peer-reviewed material.  
  
Online sources: These papers include a variety of digital resources available on the internet, such as online databases, citizen science platforms, or social media. They may contain ecological data, maps, photographs, or other multimedia content.
  
We also add fields concerning source topics and which type of information about snails is given:

- source categories: `empirical`, `review`, `theoretical`, `methodological`, `applied`, `historical`, `data_paper`, `species_checklist`, `unpublished`, `grey`, `online`
- topic: `systematics`, `genetics`, `ecology`, `biogeography`
- snail characteristics: `fossil`, `recent`, `land`, `freshwater`, `shell_morphology`, `external_morphology`, `anatomy`, `figured`, `distribution`

The final column `taxa` contains the taxa from Gran Canaria mentioned in each paper. Each taxon from the same paper is separated by the pipe symbol |.

## Task 2.4: Write comprehensive literature review {#sec-task24}

We read every source and write out a Quarto Document where we write paragraphs about source information, research objective, problem or gap addressed, findings and conclusions, limitations or weaknesses, implications or suggestions future research and how our research is related to the paper. We can also list for each source the True fields and the taxa from the dataset from @sec-task23.  
This and the previous task can be performed parallel to each other and also provide feedback to each other.  
This review can be published on the website.

## Task 2.5: Write literature report {#sec-task25}

Finally, we create a concise but complete literature report for publication. It is a summary of the previous task and contains graphs (e.g. for visualizing trends in literature topics and frequencies) and tables [cf. @groh2012bibliography]. It will also contain a list of all the encountered taxa with an indication if the name is still valid or not.

# WP3: Data management {#sec-WP3}

The data for this research project will include field survey data (location, date, species name ...) and pictures of livings snails, empty shells and locations. All data is backed up to the cloud by syncing with Google Drive.

## Task 3.1: Create picture database {#sec-task31}

We store pictures in three folders:

- living snails
- empty shells
- pictures of survey locations and habitats

A Google sheet is made in each folder which links pictures to meta data on precise location, time, remarks ...

## Task 3.2: Create occurrence database {#sec-task32}

identification etc.

## Task 3.3: Data quality and validation {#sec-task33}

https://cran.r-project.org/web/packages/validate/vignettes/cookbook.html
https://rich-iannone.github.io/pointblank/

# WP4: Data processing and synthesis {#sec-WP4}

## Task 4.1: Use version control for reproducibility {#sec-task41}

## Task 4.2: {#sec-task42}

## Task 4.3: {#sec-task43}

## Task 4.4: {#sec-task44}

## Task 4.5: {#sec-task45}

## Task 4.6: {#sec-task46}

## Task 4.7: {#sec-task47}

## Task 4.8: {#sec-task48}


# References {#sec-references}
